---
title: "Gene annotation tracks for genomic plots in R"
description: >
  Add scored, strand-aware gene-body annotations to interactive rainfall and
  GISTIC plots with MutGlyph.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene annotation tracks for genomic plots in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Gene annotation tracks add genomic landmarks and local context to interactive
plots. At a wide view, familiar symbols help with navigation; at a close view,
gene bodies show which genes are near a mutation cluster, copy-number peak, or
other feature of interest.

MutGlyph uses the same generic `annotationTracks` argument for rainfall and
GISTIC plots. The package includes an offline scored gene-body track for hg18,
hg19, and hg38.

## Use annotations with rainfall plots

This example opens at a mutation-dense region and places gene bodies below the
rainfall panel.

```{r rainfall-data}
library(MutGlyph)

genes <- mutglyph_gene_annotations("hg19")

brca <- maftools::read.maf(
  maf = system.file("extdata", "brca.maf.gz", package = "maftools"),
  verbose = FALSE
)
```

```{r rainfall-genes}
rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  region = "chr8:98,000,000-98,500,000",
  annotationTracks = list(genes = genes),
  height = 450
)
```

## Use annotations with GISTIC plots

The same annotation interface adds genomic context below amplification and
deletion profiles.

```{r gistic-data}
gistic <- maftools::readGistic(
  gisticDir = system.file("extdata", package = "maftools"),
  isTCGA = TRUE,
  verbose = FALSE
)
```

```{r gistic-genes}
gisticChromPlot(
  gistic,
  region = "chr21:39,000,000-41,000,000",
  annotationTracks = list(genes = genes),
  height = 450
)
```

## Supply a custom annotation track

Annotation tracks are named interval tracks, not a gene-specific plot option.
They can be supplied as `GRanges` objects or data frames. A track must contain
`seqnames`, `start`, `end`, and finite numeric `score` columns. `label`,
`identifier`, and `strand` are optional; supplied strand values must be
complete and use `+` or `-`.

```{r custom-track, eval = FALSE}
custom_track <- data.frame(
  seqnames = c("chr8", "chr8"),
  start = c(98000000, 98200000),
  end = c(98001000, 98202000),
  label = c("FEATURE_A", "FEATURE_B"),
  identifier = c("a", "b"),
  strand = c("+", "-"),
  score = c(10, 1)
)

rainfallPlot(
  brca,
  region = "chr8:98,000,000-98,500,000",
  annotationTracks = list(features = custom_track)
)
```

Coordinates must use the same assembly as the plot. MutGlyph checks assembly
metadata on `GRanges` inputs before rendering.

## How the annotation track is rendered

Gene bodies are shown with strand-aware `arrow-block` marks. Symbols appear
above their bodies. The view uses at most three priority lanes, keeping the `-`
and `+` strands stable and using the remaining lane as a fallback. Exons are
omitted so the track stays compact in embedded RMarkdown widgets.

## Packaged gene annotations

`mutglyph_gene_annotations()` returns a standard `GRanges` object with:

- genomic sequence and 1-based, closed coordinates;
- strand;
- gene symbol and NCBI GeneID; and
- a numeric popularity score.

Each gene is represented by one inclusive body interval after transcript
records are collapsed by NCBI GeneID. The packaged resources are prepared and
frozen with MutGlyph, so creating a plot does not require a network connection.

The coordinates come from assembly-matched [UCSC RefSeq
`refGene`](https://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/refGene.txt.gz)
records mapped through [UCSC `ncbiRefSeqLink`](https://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/ncbiRefSeqLink.txt.gz).
Symbols come from [NCBI `gene_info`](https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Homo_sapiens.gene_info.gz).

## Popularity scores and label prioritization

The score is a popularity score. It counts unique GeneID--PubMed pairs in the
NCBI GeneRIF [`generifs_basic`](https://ftp.ncbi.nlm.nih.gov/gene/GeneRIF/generifs_basic.gz)
snapshot. When labels compete for space, higher-popularity genes are preferred.
The score is only a navigation and label-layout heuristic: it does not measure
biological importance, expression, mutation significance, or evidence quality.

This popularity-based label-scoring idea comes from the [HiGlass gene
annotation track workflow](https://docs.higlass.io/data_preparation.html#gene-annotation-tracks).
MutGlyph applies that idea to interactive label prioritization while keeping
annotation inputs and rendering composable in R.

## Inspect the packaged annotation object

The full object is large, so compact inspection is usually most useful:

```{r inspect-genes}
head(genes)
attr(genes, "mutglyph_annotation")
```

## Practical note on output size

The hg19 resource contains about 27,600 gene bodies. Including it makes the
resulting RMarkdown HTML larger because the annotation rows are embedded in
the widget specification. In a current measurement, the track added about
1.2 MB to the serialized rainfall widget payload. The exact size varies with
the plot and the number of annotation tracks.
