---
title: "Gene annotation tracks for genomic plots in R"
description: >
  Add scored, strand-aware gene-body annotations to interactive rainfall and
  GISTIC plots with MutGlyph.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene annotation tracks for genomic plots in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Gene annotation tracks provide genomic landmarks and local context. At a
wide view, familiar gene symbols help orient navigation; at a close view,
gene bodies show which genes are near a mutation cluster, copy-number peak,
or other feature of interest.

## The built-in scored gene track

MutGlyph includes an offline gene-body annotation resource for hg18, hg19,
and hg38:

```{r load-genes}
library(MutGlyph)

genes <- mutglyph_gene_annotations("hg19")
genes
```

The returned object is a `GRanges` object with the genomic sequence, 1-based
closed coordinates, strand, gene symbol, identifier, and numeric score. The
score is a label-priority heuristic: when several labels compete for space,
higher-scoring genes are preferred. It does not represent biological
importance, expression, or mutation significance.

Gene bodies are rendered with strand-aware `arrow-block` marks. The annotation
view uses at most three lanes, preferring stable lanes for the `-` and `+`
strands and using the remaining lane as a fallback. Symbols are shown above
their bodies. Exons are intentionally omitted so the track stays compact in
embedded RMarkdown widgets.

The label-scoring idea comes from the [HiGlass gene annotation track
workflow](https://docs.higlass.io/data_preparation.html#gene-annotation-tracks).
MutGlyph uses that idea for interactive label prioritization while keeping the
annotation input and rendering composable in R.

## Use annotations with rainfall plots

The same named `annotationTracks` interface works with rainfall plots. This
example opens at a mutation-dense region and shows the scored gene bodies
below the rainfall panel.

```{r rainfall-data}
brca <- maftools::read.maf(
  maf = system.file("extdata", "brca.maf.gz", package = "maftools"),
  verbose = FALSE
)
```

```{r rainfall-genes}
rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  region = "chr8:98,000,000-98,500,000",
  annotationTracks = list(genes = genes),
  height = 540
)
```

Annotation tracks are generic named interval tracks, not a rainfall-specific
gene argument. This means a plot can contain several tracks, each supplied as
a `GRanges` object or data frame.

## Use annotations with GISTIC plots

GISTIC plots use the same interface. Here the gene track provides context for
a narrow region containing amplification and deletion profiles.

```{r gistic-data}
gistic <- maftools::readGistic(
  gisticDir = system.file("extdata", package = "maftools"),
  isTCGA = TRUE,
  verbose = FALSE
)
```

```{r gistic-genes}
gisticChromPlot(
  gistic,
  region = "chr21:39,000,000-41,000,000",
  annotationTracks = list(genes = genes),
  height = 540
)
```

## Supply a custom annotation track

Custom tracks must contain `seqnames`, `start`, `end`, and finite numeric
`score` columns. `label`, `identifier`, and `strand` are optional; `strand`
must be complete when supplied and use `+` or `-` values.

```{r custom-track, eval = FALSE}
custom_track <- data.frame(
  seqnames = c("chr8", "chr8"),
  start = c(98000000, 98200000),
  end = c(98001000, 98202000),
  label = c("FEATURE_A", "FEATURE_B"),
  identifier = c("a", "b"),
  strand = c("+", "-"),
  score = c(10, 1)
)

rainfallPlot(
  brca,
  region = "chr8:98,000,000-98,500,000",
  annotationTracks = list(features = custom_track)
)
```

Coordinates must use the same assembly as the plot. `GRanges` inputs may carry
assembly metadata, which MutGlyph checks before rendering.
