---
title: "Gene annotation tracks for genomic plots in R"
description: >
  Add scored, strand-aware gene-body annotations to interactive rainfall and
  GISTIC plots with MutGlyph.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene annotation tracks for genomic plots in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Gene annotation tracks provide genomic landmarks and local context. At a
wide view, familiar gene symbols help orient navigation; at a close view,
gene bodies show which genes are near a mutation cluster, copy-number peak,
or other feature of interest.

## What is a gene annotation track?

MutGlyph includes an offline gene-body annotation resource for hg18, hg19,
and hg38. These packaged tracks are RefSeq/NCBI-derived, not GENCODE-derived:
the coordinates come from assembly-matched UCSC RefSeq `refGene` records,
which are mapped to NCBI GeneIDs through `ncbiRefSeqLink`; gene symbols come
from NCBI `gene_info`.

The returned object is a `GRanges` object with the genomic sequence, 1-based
closed coordinates, strand, gene symbol, identifier, and numeric score. Each
gene is represented by one inclusive body interval after transcript records
are collapsed by NCBI GeneID.

The score is a popularity score: it counts unique GeneID–PubMed pairs in the
NCBI GeneRIF `generifs_basic` snapshot. When several labels compete for space,
higher-popularity genes are preferred. This is a navigation and label-layout
heuristic; it does not represent biological importance, expression, mutation
significance, or evidence quality.

The source snapshots are prepared and frozen with the package rather than
downloaded when a plot is created. The upstream sources are [UCSC
RefSeq](https://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/refGene.txt.gz),
[UCSC `ncbiRefSeqLink`](https://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/ncbiRefSeqLink.txt.gz),
[NCBI `gene_info`](https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Homo_sapiens.gene_info.gz),
and [NCBI GeneRIF `generifs_basic`](https://ftp.ncbi.nlm.nih.gov/gene/GeneRIF/generifs_basic.gz).

Gene bodies are rendered with strand-aware `arrow-block` marks. The annotation
view uses at most three lanes, preferring stable lanes for the `-` and `+`
strands and using the remaining lane as a fallback. Symbols are shown above
their bodies. Exons are intentionally omitted so the track stays compact in
embedded RMarkdown widgets.

The popularity-based label-scoring idea comes from the [HiGlass gene annotation track
workflow](https://docs.higlass.io/data_preparation.html#gene-annotation-tracks).
MutGlyph uses that idea for interactive label prioritization while keeping the
annotation input and rendering composable in R.

For the current hg19 resource, the packaged track contains about 27,600 gene
bodies. Including it makes the resulting RMarkdown HTML larger because the
annotation rows are embedded in the widget specification. In a current
measurement, the hg19 track added approximately 3.2 MB to a rainfall widget's
serialized JSON (the compressed package resource itself is about 0.5 MB).
The exact size varies with the plot and the number of tracks.

The examples below show the usual workflow first. Details of the returned
`GRanges` object are at the end of the article.

## Use annotations with rainfall plots

The same named `annotationTracks` interface works with rainfall plots. This
example opens at a mutation-dense region and shows the scored gene bodies
below the rainfall panel.

```{r rainfall-data}
library(MutGlyph)

genes <- mutglyph_gene_annotations("hg19")

brca <- maftools::read.maf(
  maf = system.file("extdata", "brca.maf.gz", package = "maftools"),
  verbose = FALSE
)
```

```{r rainfall-genes}
rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  region = "chr8:98,000,000-98,500,000",
  annotationTracks = list(genes = genes),
  height = 540
)
```

Annotation tracks are generic named interval tracks, not a rainfall-specific
gene argument. This means a plot can contain several tracks, each supplied as
a `GRanges` object or data frame.

## Use annotations with GISTIC plots

GISTIC plots use the same interface. Here the gene track provides context for
a narrow region containing amplification and deletion profiles.

```{r gistic-data}
gistic <- maftools::readGistic(
  gisticDir = system.file("extdata", package = "maftools"),
  isTCGA = TRUE,
  verbose = FALSE
)
```

```{r gistic-genes}
gisticChromPlot(
  gistic,
  region = "chr21:39,000,000-41,000,000",
  annotationTracks = list(genes = genes),
  height = 540
)
```

## Supply a custom annotation track

Custom tracks must contain `seqnames`, `start`, `end`, and finite numeric
`score` columns. `label`, `identifier`, and `strand` are optional; `strand`
must be complete when supplied and use `+` or `-` values.

```{r custom-track, eval = FALSE}
custom_track <- data.frame(
  seqnames = c("chr8", "chr8"),
  start = c(98000000, 98200000),
  end = c(98001000, 98202000),
  label = c("FEATURE_A", "FEATURE_B"),
  identifier = c("a", "b"),
  strand = c("+", "-"),
  score = c(10, 1)
)

rainfallPlot(
  brca,
  region = "chr8:98,000,000-98,500,000",
  annotationTracks = list(features = custom_track)
)
```

Coordinates must use the same assembly as the plot. `GRanges` inputs may carry
assembly metadata, which MutGlyph checks before rendering.

## Inspect the packaged annotation object

The built-in resource is a standard `GRanges` object. Its ranges are one
inclusive gene-body interval per NCBI GeneID, with symbols, identifiers,
strands, and popularity scores in the metadata columns. The full object is
large, so a compact inspection is usually more useful in a vignette:

```{r inspect-genes}
head(genes)
attr(genes, "mutglyph_annotation")
```
