---
title: "Interactive protein lollipop plots in R"
description: >
  Explore protein mutations with interactive lollipop plots in R using MAF
  objects or data frames, custom domains, labels, and cohort comparisons.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive protein lollipop plots in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`lollipopPlot()` follows the familiar `maftools::lollipopPlot()` data
conventions but uses a cleaner interactive composition. Mutation events stay
at their true amino-acid positions, and domains are an ordinary data frame that
can be inspected, edited, or replaced.

As in maftools, `refSeqID` and `proteinID` select the protein-domain model, and
the longest bundled protein is used when neither is supplied. MutGlyph shows
the selected transcript and protein identifiers below the plot title. If the
mutation input contains compatible RefSeq metadata, it warns when the mutations
contain mixed isoforms or disagree with the domain model. Such mutations are
not silently removed.

```{r data}
library(MutGlyph)

laml <- maftools::read.maf(
  maf = system.file("extdata", "tcga_laml.maf.gz", package = "maftools"),
  verbose = FALSE
)
```

## Start from a MAF object

The small table below is a frozen, offline snapshot derived from InterPro's
representative-domain matches for UniProt P36888 (FLT3), retrieved 2026-08-13.
Current annotations can instead be requested explicitly with
`mutglyph_interpro_domains("P36888")`; plotting itself never requires network
access. The retrieval and selection workflow is retained under
`data-raw/protein-domains/` in the source repository.

```{r flt3-domains}
flt3_domains <- data.frame(
  start = c(246, 438, 564, 756),
  end = c(357, 531, 695, 958),
  label = c("Ig-like", "Ig-like", "Kinase N", "Kinase C"),
  description = c(
    "Immunoglobulin-like region",
    "Immunoglobulin-like region",
    "Protein kinase N-terminal region",
    "Protein kinase C-terminal region"
  ),
  accession = c(
    "G3DSA:2.60.40.10", "G3DSA:2.60.40.10",
    "G3DSA:3.30.200.20", "G3DSA:1.10.510.10"
  ),
  source_database = "cathgene3d",
  protein_id = "P36888",
  protein_length = 993
)
```

```{r lollipop-basic}
lollipopPlot(
  laml,
  gene = "FLT3",
  AACol = "Protein_Change",
  domains = flt3_domains,
  height = 350
)
```

**Try the plot:** Zoom and pan along the protein coordinate scale, and hover
over mutation markers or domains for recurrence and annotation details.

## Compare two cohorts

`lollipopPlot2()` is the interactive counterpart to
`maftools::lollipopPlot2()`. It mirrors two cohorts around one shared protein
model while scaling their recurrence axes independently. Common maftools calls
therefore need little more than a namespace change.

```{r apl-data}
primary <- maftools::read.maf(
  system.file("extdata", "APL_primary.maf.gz", package = "maftools"),
  verbose = FALSE
)
relapse <- maftools::read.maf(
  system.file("extdata", "APL_relapse.maf.gz", package = "maftools"),
  verbose = FALSE
)
```

```{r lollipop-two-cohort}
lollipopPlot2(
  m1 = primary,
  m2 = relapse,
  gene = "FLT3",
  AACol1 = "amino_acid_change",
  AACol2 = "amino_acid_change",
  m1_name = "Primary",
  m2_name = "Relapse",
  m1_label = 835,
  m2_label = 835,
  domains = flt3_domains,
  height = 350
)
```

Both stem sets extend to the center of the protein track, where the protein is
intended to cover their ends. The joins currently look somewhat ugly because
the bundled GenomeSpy 0.84 cannot z-order complete children of a vertical
concatenation: the later lower cohort draws its stems over the protein. A future
GenomeSpy version with broader `zindex` support will render the intended
layering cleanly without changing the plot geometry.

## Annotate selected mutations

`labelPos` accepts amino-acid positions or `"all"`. As in maftools, changes at
the same residue are collapsed by default: the FLT3 substitutions at residue
835 become one label such as `D835E/H/Y`, anchored above the tallest lollipop
at that position. Set `collapsePosLabel = FALSE` when each change needs its own
label.

```{r lollipop-annotations}
lollipopPlot(
  laml,
  gene = "FLT3",
  AACol = "Protein_Change",
  domains = flt3_domains,
  labelPos = c(599, 835),
  labPosSize = 0.9,
  labPosAngle = 35,
  collapsePosLabel = TRUE,
  height = 350
)
```

Dense hotspots often contain several different protein changes at the same or
nearby residues. The displaced layout reserves marker space in pixels, labels
recurrent changes, and uses curved connectors to preserve the true protein
position. Its logarithmic recurrence axis keeps moderately recurrent mutations
visible beside dominant hotspots. Singletons are omitted by default; set
`minCount = 1` to include them.

```{r lollipop-displaced}
lollipopPlot(
  laml,
  gene = "FLT3",
  AACol = "Protein_Change",
  domains = flt3_domains,
  layout = "displaced",
  count = "samples",
  height = 400
)
```

Both mutation-event and distinct-sample counts remain available in tooltips and
in the generated specification. The mutation input can also be a regular data frame containing
`position` and optional `mutation`, `variant_class`, `sample`, and `count`
columns, so custom and pre-aggregated analyses use the same renderer.

## Compose custom mutation and domain data

The bundled `pik3ca_tcga_brca` table demonstrates the same plotting path
without a MAF object. It contains 26 recurrent PIK3CA mutations from the
open-access GDC TCGA-BRCA masked somatic mutation data, pre-aggregated by
distinct tumor sample. Its preparation is an R script under `data-raw/` in the
source repository.

> **TCGA acknowledgement.** The results shown here are in whole or part based
> upon data generated by the TCGA Research Network:
> <https://www.cancer.gov/tcga>.

The five-row domain table is deliberately inline: domains are just data and can
be edited or replaced. These regions are a frozen snapshot of the main domain
features in reviewed UniProt entry P42336.
The same provenance workflow records how these features were selected.

```{r pik3ca-data}
data(pik3ca_tcga_brca)

pik3ca_domains <- data.frame(
  start = c(16, 187, 330, 517, 765),
  end = c(105, 289, 487, 694, 1051),
  label = c("ABD", "RBD", "C2", "Helical", "Kinase"),
  description = c(
    "PI3K adaptor-binding domain",
    "PI3K Ras-binding domain",
    "C2 PI3K-type domain",
    "PIK helical domain",
    "PI3K/PI4K catalytic domain"
  ),
  protein_id = "P42336",
  protein_length = 1068
)
```

```{r lollipop-custom-data}
lollipopPlot(
  data = pik3ca_tcga_brca,
  gene = "PIK3CA",
  domains = pik3ca_domains,
  count = "samples",
  layout = "displaced",
  height = 420
)
```
