---
title: "Introduction to MutGlyph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MutGlyph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

MutGlyph builds interactive cancer-genomics plots from
[`maftools`](https://bioconductor.org/packages/maftools/) MAF objects and renders
them with [GenomeSpy](https://genomespy.app/). Oncoplots, protein lollipops, and
rainfall plots support tooltips, zooming, and SVG export.

`oncoplot()` deliberately imitates `maftools::oncoplot()`: it accepts
the same MAF objects and follows familiar argument names, ordering behavior,
and plot composition. It is therefore an almost drop-in replacement in common
workflows, often requiring only a namespace change. MutGlyph does not
attempt to reproduce every maftools option.

## Start with a familiar oncoplot

The bundled TCGA acute myeloid leukemia data makes a convenient small example.
This starting point deliberately corresponds to the [basic top-ten
oncoplot](https://www.bioconductor.org/packages/release/bioc/vignettes/maftools/inst/doc/maftools.html#721_Drawing_oncoplots)
in the maftools introduction, but uses MutGlyph for rendering.

```{r data}
library(MutGlyph)

laml <- maftools::read.maf(
  maf = system.file("extdata", "tcga_laml.maf.gz", package = "maftools"),
  clinicalData = system.file(
    "extdata",
    "tcga_laml_annot.tsv",
    package = "maftools"
  ),
  verbose = FALSE
)
```

```{r basic-oncoplot}
oncoplot(maf = laml, top = 10, height = 500)
```

The columns share one sample scale. Scroll or pinch over the plot to zoom and
pan, and hover over marks for details. Sample labels can be enabled even for a
dense cohort: ranged text keeps them hidden until the columns are wide enough.
The **Save as SVG** button exports the current composition.

## Continue with the plot guides

- [Customize oncoplots](oncoplots.html) with clinical tracks, TiTv, summary
  bars, colors, selection, and GISTIC calls.
- [Map mutations onto proteins](lollipop-plots.html) using MAF objects or
  ordinary data frames and composable domain annotations.
- [Explore mutation clustering](rainfall-plots.html) with rainfall plots and
  optional kataegis detection.

## Use the generated specification

MutGlyph returns an htmlwidget containing the exact GenomeSpy specification
used in the browser. `as_json()` creates a portable copy for inspection,
versioning, or experimentation in the GenomeSpy Playground.

```{r specification}
plot <- oncoplot(laml, top = 10, height = 500)
spec_json <- as_json(plot, pretty = TRUE)
cat(substr(spec_json, 1, 160), "...\n")
```

Widget width is responsive by default. Set `height` when embedding the plot in
a constrained layout; matrix height is controlled independently with
`rowHeight`.

## Data acknowledgement

The results shown in these articles are in whole or part based upon data
generated by the TCGA Research Network: <https://www.cancer.gov/tcga>.

```{r session-info}
sessionInfo()
```
