---
title: "Introduction to MutGlyph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MutGlyph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

MutGlyph builds interactive cancer-genomics plots from
[`maftools`](https://bioconductor.org/packages/maftools/) MAF objects and renders
them with [GenomeSpy](https://genomespy.app/). Oncoplots, protein lollipops, and
rainfall plots support tooltips, zooming, and SVG export.

`mutglyph_oncoplot()` deliberately imitates `maftools::oncoplot()`: it accepts
the same MAF objects and follows familiar argument names, ordering behavior,
and plot composition. It is therefore an almost drop-in replacement in common
workflows, often requiring only a change of function name. MutGlyph does not
attempt to reproduce every maftools option.

## A basic oncoplot

The bundled TCGA acute myeloid leukemia data makes a convenient small example.
This starting point deliberately corresponds to the [basic top-ten
oncoplot](https://www.bioconductor.org/packages/release/bioc/vignettes/maftools/inst/doc/maftools.html#721_Drawing_oncoplots)
in the maftools introduction, but uses MutGlyph for rendering.

```{r data}
library(MutGlyph)

laml <- maftools::read.maf(
  maf = system.file("extdata", "tcga_laml.maf.gz", package = "maftools"),
  clinicalData = system.file(
    "extdata",
    "tcga_laml_annot.tsv",
    package = "maftools"
  ),
  verbose = FALSE
)
```

```{r basic-oncoplot}
mutglyph_oncoplot(maf = laml, top = 10)
```

The columns share one sample scale. Scroll or pinch over the plot to zoom and
pan, and hover over marks for details. Sample labels can be enabled even for a
dense cohort: ranged text keeps them hidden until the columns are wide enough.
The **Save as SVG** button exports the current composition.

## Choose genes and samples

Use `top` for the most frequently altered genes, or provide an explicit gene
list. `keepGeneOrder = TRUE` preserves the supplied order. The same pattern
works for a selected and ordered set of samples.

```{r selection, eval = FALSE}
mutglyph_oncoplot(
  laml,
  genes = c("NPM1", "FLT3", "DNMT3A", "IDH1", "IDH2"),
  keepGeneOrder = TRUE,
  sampleOrder = c("TCGA-AB-2945", "TCGA-AB-2965"),
  removeNonMutated = TRUE
)
```

`minMut` provides a concise alternative when the exact genes are not known. A
value below one is treated as a cohort fraction; a value of one or more is a
sample count.

```{r threshold, eval = FALSE}
mutglyph_oncoplot(laml, minMut = 0.05, genesToIgnore = "TTN")
```

## Add clinical and sequence context

Clinical tracks can be categorical or numeric. Each track gets an independent
scale, and categorical tracks use GenomeSpy's default categorical palette
unless a Vega scheme or explicit mapping is supplied. The transition/
transversion track is computed from the MAF variants.

```{r annotated-oncoplot}
mutglyph_oncoplot(
  laml,
  top = 10,
  rowHeight = 28,
  clinicalFeatures = c("FAB_classification", "days_to_last_followup"),
  annotationColor = list(
    days_to_last_followup = "blues"
  ),
  sortByAnnotation = TRUE,
  annotationOrder = list(
    FAB_classification = c("M5", "M4", "M2")
  ),
  draw_titv = TRUE,
  showTumorSampleBarcodes = TRUE,
  titleText = "TCGA acute myeloid leukemia"
)
```

Mutation-class colors are partial overrides, so a plot can emphasize a few
classes without redefining the whole palette.

```{r mutation-colors, eval = FALSE}
mutglyph_oncoplot(
  laml,
  top = 10,
  colors = c(
    Missense_Mutation = "#00897B",
    Multi_Hit = "#D81B60"
  )
)
```

## Replace or extend summary bars

Custom bars use a two-column data frame. The first column identifies a sample
or gene; the numeric second column supplies the value and its name becomes the
axis title. A numeric clinical field name can be used directly for the top bar.

```{r summary-data}
genes <- as.character(maftools::getGeneSummary(laml)$Hugo_Symbol[1:10])

variants <- maftools::subsetMaf(
  laml,
  genes = genes,
  fields = c("Hugo_Symbol", "i_TumorVAF_WU"),
  includeSyn = FALSE,
  mafObj = FALSE
)

mean_vaf <- aggregate(i_TumorVAF_WU ~ Hugo_Symbol, variants, mean)
names(mean_vaf) <- c("gene", "Mean VAF (%)")
```

```{r summary-bars}
mutglyph_oncoplot(
  laml,
  genes = genes,
  keepGeneOrder = TRUE,
  topBarData = "days_to_last_followup",
  leftBarData = mean_vaf,
  leftBarLims = c(0, 100),
  titleText = "Follow-up and mean variant allele frequency"
)
```

Custom top and right data replace their default stacked summaries. A custom
left bar adds a new gene-aligned column. Missing displayed keys are shown as
zero and reported with a warning.

## Include GISTIC calls

When `maftools::read.maf()` is given GISTIC results, MutGlyph adds gene-level
`Amp` and `Del` events to the matrix. Copy-number events occupy half of a gene
band so that a sequence mutation in the same cell remains visible. The default
top bars include both sequence mutations and copy-number events.

```{r gistic}
extdata <- system.file("extdata", package = "maftools")

laml_gistic <- maftools::read.maf(
  maf = file.path(extdata, "tcga_laml.maf.gz"),
  clinicalData = file.path(extdata, "tcga_laml_annot.tsv"),
  gisticAllLesionsFile = file.path(extdata, "all_lesions.conf_99.txt"),
  gisticAmpGenesFile = file.path(extdata, "amp_genes.conf_99.txt"),
  gisticDelGenesFile = file.path(extdata, "del_genes.conf_99.txt"),
  gisticScoresFile = file.path(extdata, "scores.gistic"),
  verbose = FALSE
)

mutglyph_oncoplot(
  laml_gistic,
  top = 10,
  titleText = "Sequence mutations and GISTIC calls"
)
```

Set `includeColBarCN = FALSE` to keep the copy-number layer in the matrix but
limit the top bars to sequence mutations.

## Map mutations onto a protein

`mutglyph_lollipop_plot()` follows the familiar `maftools::lollipopPlot()` data
conventions but uses a cleaner interactive composition. Mutation events stay
at their true amino-acid positions, and domains are an ordinary data frame that
can be inspected, edited, or replaced.

The small table below is a frozen, offline snapshot derived from InterPro's
representative-domain matches for UniProt P36888 (FLT3), retrieved 2026-08-13.
Current annotations can instead be requested explicitly with
`mutglyph_interpro_domains("P36888")`; plotting itself never requires network
access.

```{r flt3-domains}
flt3_domains <- data.frame(
  start = c(246, 438, 564, 756),
  end = c(357, 531, 695, 958),
  label = c("Ig-like", "Ig-like", "Kinase N", "Kinase C"),
  description = c(
    "Immunoglobulin-like region",
    "Immunoglobulin-like region",
    "Protein kinase N-terminal region",
    "Protein kinase C-terminal region"
  ),
  accession = c(
    "G3DSA:2.60.40.10", "G3DSA:2.60.40.10",
    "G3DSA:3.30.200.20", "G3DSA:1.10.510.10"
  ),
  source_database = "cathgene3d",
  protein_id = "P36888",
  protein_length = 993
)
```

```{r lollipop-basic}
mutglyph_lollipop_plot(
  laml,
  gene = "FLT3",
  AACol = "Protein_Change",
  domains = flt3_domains,
  height = 420
)
```

Dense hotspots often contain several different protein changes at the same or
nearby residues. The displaced layout reserves marker space in pixels, labels
recurrent changes, and uses curved connectors to preserve the true protein
position. Its logarithmic recurrence axis keeps moderately recurrent mutations
visible beside dominant hotspots. Singletons are omitted by default; set
`minCount = 1` to include them.

```{r lollipop-displaced}
mutglyph_lollipop_plot(
  laml,
  gene = "FLT3",
  AACol = "Protein_Change",
  domains = flt3_domains,
  layout = "displaced",
  count = "samples",
  height = 500
)
```

Set `labelPos` to selected residues in the basic view. Both mutation-event and
distinct-sample counts remain available in tooltips and in the generated
specification. The mutation input can also be a regular data frame containing
`position` and optional `mutation`, `variant_class`, `sample`, and `count`
columns, so custom and pre-aggregated analyses use the same renderer.

### Compose custom mutation and domain data

The bundled `pik3ca_tcga_brca` table demonstrates the same plotting path
without a MAF object. It contains 26 recurrent PIK3CA mutations from the
open-access GDC TCGA-BRCA masked somatic mutation data, pre-aggregated by
distinct tumor sample. Its preparation is an R script under `data-raw/` in the
source repository.

> **TCGA acknowledgement.** The results shown here are in whole or part based
> upon data generated by the TCGA Research Network:
> <https://www.cancer.gov/tcga>.

The five-row domain table is deliberately inline: domains are just data and can
be edited or replaced. These regions are a frozen snapshot of the main domain
features in reviewed UniProt entry P42336.

```{r pik3ca-data}
data(pik3ca_tcga_brca)

pik3ca_domains <- data.frame(
  start = c(16, 187, 330, 517, 765),
  end = c(105, 289, 487, 694, 1051),
  label = c("ABD", "RBD", "C2", "Helical", "Kinase"),
  description = c(
    "PI3K adaptor-binding domain",
    "PI3K Ras-binding domain",
    "C2 PI3K-type domain",
    "PIK helical domain",
    "PI3K/PI4K catalytic domain"
  ),
  protein_id = "P42336",
  protein_length = 1068
)
```

```{r lollipop-custom-data}
mutglyph_lollipop_plot(
  pik3ca_tcga_brca,
  gene = "PIK3CA",
  domains = pik3ca_domains,
  count = "samples",
  layout = "displaced",
  height = 500
)
```

## Explore mutation clustering with a rainfall plot

`mutglyph_rainfall_plot()` is the interactive counterpart of
`maftools::rainfallPlot()`. It defaults to the most mutated sample and plots
chromosome-local inter-event distances for SNPs, including synonymous
variants. Scroll or pinch to inspect dense genomic regions.

The bundled BRCA example contains several potential kataegis loci. With
`detectChangePoints = TRUE`, MutGlyph applies the same operational criterion as
maftools—at least six consecutive mutations with an average intermutation
distance no greater than 1,000 bp. Arrows point to the clusters and subtle
shading shows their complete genomic spans.

```{r rainfall}
brca <- maftools::read.maf(
  maf = system.file("extdata", "brca.maf.gz", package = "maftools"),
  verbose = FALSE
)

mutglyph_rainfall_plot(
  brca,
  detectChangePoints = TRUE,
  pointSize = 0.5,
  height = 360
)
```

Detected loci remain available in the returned specification without writing a
TSV file:

```{r rainfall-data, eval = FALSE}
rainfall <- mutglyph_rainfall_plot(brca, detectChangePoints = TRUE)
rainfall$x$spec$datasets$kataegis
```

## Use the generated specification

MutGlyph returns an htmlwidget containing the exact GenomeSpy specification
used in the browser. `as_json()` creates a portable copy for inspection,
versioning, or experimentation in the GenomeSpy Playground.

```{r specification}
plot <- mutglyph_oncoplot(laml, top = 10)
spec_json <- as_json(plot, pretty = TRUE)
cat(substr(spec_json, 1, 160), "...\n")
```

Widget width is responsive by default. Set `height` when embedding the plot in
a constrained layout; matrix height is controlled independently with
`rowHeight`.

```{r session-info}
sessionInfo()
```
