---
title: "Get started with interactive cancer genomics in R"
description: >
  Create an interactive oncoplot in R from a maftools MAF object and explore
  the zooming, panning, tooltips, and export controls provided by MutGlyph.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with interactive cancer genomics in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This page takes an existing MAF object to a first interactive MutGlyph plot.
MutGlyph is exclusively a visualization package: use maftools or another
analysis workflow to prepare and analyze variants, then use MutGlyph to render
the results with an embedded
[GenomeSpy interactive genomic visualization](https://genomespy.app/).

## Start with a familiar oncoplot

The TCGA acute myeloid leukemia data bundled with maftools makes a convenient
small example.
This starting point deliberately corresponds to the [basic top-ten
oncoplot](https://www.bioconductor.org/packages/release/bioc/vignettes/maftools/inst/doc/maftools.html#721_Drawing_oncoplots)
in the maftools introduction, but uses MutGlyph for rendering.

```{r data}
laml <- maftools::read.maf(
  maf = system.file("extdata", "tcga_laml.maf.gz", package = "maftools"),
  clinicalData = system.file(
    "extdata",
    "tcga_laml_annot.tsv",
    package = "maftools"
  ),
  verbose = FALSE
)
```

```{r basic-oncoplot}
MutGlyph::oncoplot(maf = laml, top = 10, height = 500)
```

The columns share one sample scale. Scroll or pinch over the plot to zoom and
pan, and hover over marks for details. Sample labels can be enabled even for a
dense cohort: ranged text keeps them hidden until the columns are wide enough.
Hover over the widget to reveal controls for fullscreen viewing, PNG or SVG
image export, and downloading the generated GenomeSpy specification.

## Continue with the plot guides

- [Oncoplots](oncoplots.html) covers clinical tracks, TiTv, summary
  bars, colors, selection, and GISTIC calls.
- [Map mutations onto proteins](lollipop-plots.html) using MAF objects or
  ordinary data frames and composable domain annotations.
- [Explore mutation clustering](rainfall-plots.html) with rainfall plots and
  optional kataegis detection.
- [Browse GISTIC copy-number landscapes](gistic-plots.html) with mirrored
  amplification and deletion profiles.

## Use the generated specification

MutGlyph returns an htmlwidget containing the exact GenomeSpy specification
used in the browser. `as_json()` creates a portable copy for inspection,
versioning, or experimentation in the GenomeSpy Playground.

```{r specification}
plot <- MutGlyph::oncoplot(laml, top = 10, height = 500)
spec_json <- MutGlyph::as_json(plot, pretty = TRUE)
cat(substr(spec_json, 1, 160), "...\n")
```

Widget width is responsive by default. Set `height` when embedding the plot in
a constrained layout; matrix height is controlled independently with
`rowHeight`.

## Data acknowledgement

The results shown in these articles are in whole or part based upon data
generated by the TCGA Research Network: <https://www.cancer.gov/tcga>.

```{r session-info}
sessionInfo()
```
