---
title: "Customizing oncoplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing oncoplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This article builds on the [basic oncoplot](MutGlyph.html) and covers the main
ways to select, annotate, and summarize a cohort. The examples use the bundled
TCGA acute myeloid leukemia data.

```{r data}
library(MutGlyph)

laml <- maftools::read.maf(
  maf = system.file("extdata", "tcga_laml.maf.gz", package = "maftools"),
  clinicalData = system.file(
    "extdata",
    "tcga_laml_annot.tsv",
    package = "maftools"
  ),
  verbose = FALSE
)
```

## Choose genes and samples

Use `top` for the most frequently altered genes, or provide an explicit gene
list. `keepGeneOrder = TRUE` preserves the supplied order. The same pattern
works for a selected and ordered set of samples.

```{r selection, eval = FALSE}
oncoplot(
  laml,
  genes = c("NPM1", "FLT3", "DNMT3A", "IDH1", "IDH2"),
  keepGeneOrder = TRUE,
  sampleOrder = c("TCGA-AB-2945", "TCGA-AB-2965"),
  removeNonMutated = TRUE,
  height = 500
)
```

`minMut` provides a concise alternative when the exact genes are not known. A
value below one is treated as a cohort fraction; a value of one or more is a
sample count.

```{r threshold, eval = FALSE}
oncoplot(laml, minMut = 0.05, genesToIgnore = "TTN", height = 500)
```

## Add clinical and sequence context

Clinical tracks can be categorical or numeric. Each track gets an independent
scale, and categorical tracks use GenomeSpy's default categorical palette
unless a Vega scheme or explicit mapping is supplied. The transition/
transversion track is computed from the MAF variants.

```{r annotated-oncoplot}
oncoplot(
  laml,
  top = 10,
  rowHeight = 28,
  clinicalFeatures = c("FAB_classification", "days_to_last_followup"),
  annotationColor = list(
    days_to_last_followup = "blues"
  ),
  sortByAnnotation = TRUE,
  annotationOrder = list(
    FAB_classification = c("M5", "M4", "M2")
  ),
  draw_titv = TRUE,
  showTumorSampleBarcodes = TRUE,
  titleText = "TCGA acute myeloid leukemia",
  height = 700
)
```

Mutation-class colors are partial overrides, so a plot can emphasize a few
classes without redefining the whole palette.

```{r mutation-colors, eval = FALSE}
oncoplot(
  laml,
  top = 10,
  colors = c(
    Missense_Mutation = "#00897B",
    Multi_Hit = "#D81B60"
  ),
  height = 500
)
```

## Replace or extend summary bars

Custom bars use a two-column data frame. The first column identifies a sample
or gene; the numeric second column supplies the value and its name becomes the
axis title. A numeric clinical field name can be used directly for the top bar.

```{r summary-data}
genes <- as.character(maftools::getGeneSummary(laml)$Hugo_Symbol[1:10])

variants <- maftools::subsetMaf(
  laml,
  genes = genes,
  fields = c("Hugo_Symbol", "i_TumorVAF_WU"),
  includeSyn = FALSE,
  mafObj = FALSE
)

mean_vaf <- aggregate(i_TumorVAF_WU ~ Hugo_Symbol, variants, mean)
names(mean_vaf) <- c("gene", "Mean VAF (%)")
```

```{r summary-bars}
oncoplot(
  laml,
  genes = genes,
  keepGeneOrder = TRUE,
  topBarData = "days_to_last_followup",
  leftBarData = mean_vaf,
  leftBarLims = c(0, 100),
  titleText = "Follow-up and mean variant allele frequency",
  height = 500
)
```

Custom top and right data replace their default stacked summaries. A custom
left bar adds a new gene-aligned column. Missing displayed keys are shown as
zero and reported with a warning.

## Include GISTIC calls

When `maftools::read.maf()` is given GISTIC results, MutGlyph adds gene-level
`Amp` and `Del` events to the matrix. Copy-number events occupy half of a gene
band so that a sequence mutation in the same cell remains visible. The default
top bars include both sequence mutations and copy-number events.

```{r gistic}
extdata <- system.file("extdata", package = "maftools")

laml_gistic <- maftools::read.maf(
  maf = file.path(extdata, "tcga_laml.maf.gz"),
  clinicalData = file.path(extdata, "tcga_laml_annot.tsv"),
  gisticAllLesionsFile = file.path(extdata, "all_lesions.conf_99.txt"),
  gisticAmpGenesFile = file.path(extdata, "amp_genes.conf_99.txt"),
  gisticDelGenesFile = file.path(extdata, "del_genes.conf_99.txt"),
  gisticScoresFile = file.path(extdata, "scores.gistic"),
  verbose = FALSE
)

oncoplot(
  laml_gistic,
  top = 10,
  titleText = "Sequence mutations and GISTIC calls",
  height = 500
)
```

Set `includeColBarCN = FALSE` to keep the copy-number layer in the matrix but
limit the top bars to sequence mutations.
