---
title: "Rainfall plots and kataegis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rainfall plots and kataegis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`rainfallPlot()` is the interactive counterpart of
`maftools::rainfallPlot()`. It defaults to the most mutated sample and plots
chromosome-local inter-event distances for SNPs, including synonymous
variants. Scroll or pinch to inspect dense genomic regions.

The bundled BRCA example contains several potential kataegis loci. With
`detectChangePoints = TRUE`, MutGlyph applies the same operational criterion as
maftools—at least six consecutive mutations with an average intermutation
distance no greater than 1,000 bp. Arrows point to the clusters and subtle
shading shows their complete genomic spans.

```{r data}
library(MutGlyph)

brca <- maftools::read.maf(
  maf = system.file("extdata", "brca.maf.gz", package = "maftools"),
  verbose = FALSE
)
```

```{r rainfall}
rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  pointSize = 0.5,
  height = 360
)
```

Detected loci remain available in the returned specification without writing a
TSV file:

```{r rainfall-data, eval = FALSE}
rainfall <- rainfallPlot(brca, detectChangePoints = TRUE)
rainfall$x$spec$datasets$kataegis
```

Pass `region` to open the interactive plot directly at a locus of interest.
Coordinate separators are optional, and the plot remains zoomable.

```{r rainfall-region}
rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  region = "chr8:98,000,000-98,500,000",
  height = 360
)
```
