---
title: "Interactive, zoomable rainfall plots and kataegis in R"
description: >
  Explore mutation spacing with interactive rainfall plots in R, zoom into
  dense genomic regions, and highlight candidate kataegis loci.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive, zoomable rainfall plots and kataegis in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`rainfallPlot()` is the interactive counterpart of
`maftools::rainfallPlot()`. It defaults to the most mutated sample and plots
chromosome-local inter-event distances for SNPs, including synonymous
variants. The zoomable genomic scale can move from a whole-genome overview into
dense mutation clusters without generating a new plot.

The bundled BRCA example contains several potential kataegis loci. With
`detectChangePoints = TRUE`, MutGlyph applies the same operational criterion as
maftools—at least six consecutive mutations with an average intermutation
distance no greater than 1,000 bp. Arrows point to the clusters and subtle
shading shows their complete genomic spans.

```{r data}
library(MutGlyph)

brca <- maftools::read.maf(
  maf = system.file("extdata", "brca.maf.gz", package = "maftools"),
  verbose = FALSE
)
```

```{r rainfall}
rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  pointSize = 0.5,
  height = 360
)
```

**Try the plot:** Scroll or pinch to zoom, drag to pan across the genome, and
hover over a mutation for its coordinates and intermutation distance.

Detected loci remain available in the returned specification without writing a
TSV file:

```{r rainfall-data, eval = FALSE}
rainfall <- rainfallPlot(brca, detectChangePoints = TRUE)
rainfall$x$spec$datasets$kataegis
```

Pass `region` to open the interactive plot directly at a locus of interest.
Coordinate separators are optional, and the plot remains zoomable.

```{r rainfall-region}
rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  region = "chr8:98,000,000-98,500,000",
  height = 360
)
```

## Add genomic context

Annotation tracks are generic named interval tracks rather than a special
rainfall-only gene argument. MutGlyph currently ships an offline scored
gene-body track for hg18, hg19, and hg38; the score prioritizes labels when
space is limited, while every gene body becomes visible when zoomed in.
Gene bodies use strand-aware `arrow-block` marks and intentionally omit exons
to keep embedded RMarkdown widgets compact.

```{r rainfall-genes}
genes <- mutglyph_gene_annotations("hg19")

rainfallPlot(
  brca,
  detectChangePoints = TRUE,
  region = "chr8:98,000,000-98,500,000",
  annotationTracks = list(genes = genes),
  height = 540
)
```

The label-priority idea is inspired by the gene annotation workflow in
[HiGlass](https://docs.higlass.io/data_preparation.html#gene-annotation-tracks).
The public argument accepts additional named `GRanges` or data-frame tracks
with `seqnames`, `start`, `end`, and finite `score` columns, plus optional
`label`, `identifier`, and `strand` columns.
